---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# supercells

  <!-- badges: start -->
  [![R-CMD-check](https://github.com/Nowosad/supercells/workflows/R-CMD-check/badge.svg)](https://github.com/Nowosad/supercells/actions)
  [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
  [![Codecov test coverage](https://codecov.io/gh/Nowosad/supercells/branch/master/graph/badge.svg)](https://codecov.io/gh/Nowosad/supercells?branch=master)
  <!-- badges: end -->

The goal of **supercells** is to utilize the concept of superpixels to a variety of spatial data. 

## Installation

<!-- You can install the released version of supercells from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("supercells") -->
<!-- ``` -->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("Nowosad/spDataLarge")
remotes::install_github("Nowosad/supercells")
```
## Example

```{r}
library(supercells)
library(terra)
library(sf)
vol = rast(system.file("raster/volcano.tif", package = "supercells"))
plot(vol)
```

```{r}
vol_slic1 = supercells(vol, k = 50, compactness = 1)
plot(vol)
plot(st_geometry(vol_slic1), add = TRUE, lwd = 0.2)
```

```{r}
vol_slic2 = supercells(vol, k = 50, compactness = 1, dist_fun = "euclidean", 
                        clean = FALSE, iter = 10)
plot(vol)
plot(st_geometry(vol_slic2), add = TRUE, lwd = 0.2)
```

## Contribution

Contributions to this package are welcome - let me know if you need other distance measures or transformations, have any suggestions, or spotted a bug. 
The preferred method of contribution is through a GitHub pull request. 
Feel also free to contact us by creating [an issue](https://github.com/nowosad/supercells/issues).
