---
title: "Experimental features of the supercells package"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental features of the supercells package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r system.file('refs.bib', package = 'supercells')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5#,
  # fig.height = 5
)
```

Superpixels is a collection of segmentation concepts of grouping pixels with similar characteristics.
It is often used in computer vision to delineate parts of RGB images that are more meaningful and easier to analyze.
When applied to RGB images, each superpixel contains similar colors that also could represent real-world objects.
A large number of methods for creating superpixels were developed in the last decades, with the SLIC algorithm (Achanta et al. (2012), <doi:10.1109/TPAMI.2012.120>) being the most prominent.

The **supercells** package aims to utilize the concept of superpixels to a variety of spatial data. 
This package works on spatial data with one variable (e.g., continuous raster), many variables (e.g., RGB rasters), and spatial patterns (e.g., areas in categorical rasters).
Therefore, it enables not only to find areas that look similar on an RGB (satellite) image, but also to regionalize areas with comparable values of one or more variables.

This vignette shows some experimental features of the **supercells** package.
To reproduce the following results on your own computer, install and attach the packages:

```{r, message=FALSE}
library(supercells)    # superpixels for spatial data
library(terra)         # spatial raster data reading and handling
library(sf)            # spatial vector data reading and handling
```

```{r}
ortho = rast(system.file("raster/ortho.tif", package = "supercells"))
```

```{r}
plot(ortho)
```

# Large data support

```{r}
sc_ortho = supercells(ortho, k = 100, compactness = 1, chunk = 150)
```

```{r}
plot(sc_ortho)
```

# Parallel calculations

```{r}
library(future)
plan(multisession, workers = 3)
sc_ortho2 = supercells(ortho, k = 100, compactness = 1, 
                       chunk = 150, future = TRUE)
```

```{r}
plot(sc_ortho2)
```

# Custom cluster centers

```{r}
set.seed(2021-11-21)
custom_centers = sf::st_as_sfc(sf::st_bbox(ortho))
custom_centers = sf::st_sample(custom_centers, 100, type = "random")
custom_centers = sf::st_sf(geom = custom_centers)
```

```{r}
plot(ortho)
plot(st_geometry(custom_centers), add = TRUE)
```

```{r}
ortho_slic1 = supercells(ortho, k = custom_centers, 
                       step = 10, compactness = 1,
                       clean = FALSE)
```

```{r}
plot(ortho)
plot(st_geometry(ortho_slic1), add = TRUE, border = "red")
```


```{r}
ortho_slic2 = supercells(ortho, k = custom_centers, 
                       step = 20, compactness = 1,
                       clean = FALSE)
```

```{r}
plot(ortho)
plot(st_geometry(ortho_slic2), add = TRUE, border = "red")
```


```{r}
ortho_slic3 = supercells(ortho, k = custom_centers, 
                       step = 20, compactness = 1,
                       clean = TRUE, minarea = 8)
```

```{r}
plot(ortho)
plot(st_geometry(ortho_slic3), add = TRUE, border = "red")
```
