---
title: "Supercells of a single raster layer"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supercells of a single raster layer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r system.file('refs.bib', package = 'supercells')`"
nocite: '@achanta2012slic'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 6.5
)
```

Superpixels are a collection of segmentation concepts of grouping pixels with similar characteristics.
In this package, we refer to them as supercells.
It is often used in computer vision to delineate parts of RGB images that are more meaningful and easier to analyze.
When applied to RGB images, each superpixel contains similar colors that also could represent real-world objects.
A large number of methods for creating superpixels were developed in the last decades, with the SLIC algorithm (Achanta et al. (2012), <doi:10.1109/TPAMI.2012.120>) being the most prominent.

The **supercells** package aims to utilize the concept of supercells for a variety of spatial data. 
This package works on spatial data with one variable (e.g., continuous raster), many variables (e.g., RGB rasters), and spatial patterns (e.g., areas in categorical rasters).
Therefore, it enables not only to find areas that look similar on an RGB (satellite) image, but also to regionalize areas with comparable values of one or more variables.

This vignette shows how to use the **supercells** package on a single raster layer.
To reproduce the following results on your own computer, install and attach the packages:

```{r, message=FALSE}
library(supercells)    # supercells for spatial data
library(terra)         # spatial raster data reading and handling
library(sf)            # spatial vector data reading and handling
```

# Default arguments

The first step is to read the input data.
Here, we are going to use `volcano.tif` example data that is already included in the **supercells** package.

```{r}
vol = rast(system.file("raster/volcano.tif", package = "supercells"))
plot(vol)
```

It contains a digital elevation model for the Maunga Whau volcano, where each value represents an elevation above sea level.

The **supercells** package can be used here to delineate areas with similar elevations.
It has one main function, `supercells()`, which expects at least three arguments:

- `x` - an input raster (an object of class SpatRaster from the **terra** package)
- `k` - an expected number of supercells
- `compactness` - a compactness value. Larger values create more compact/even (square) supercells

```{r, results='hide'}
vol_slic1 = supercells(vol, k = 50, compactness = 1)
vol_slic1
```

The `supercells()` output is an `sf` object, where each row represents a single superpixel.
It stores supercell ids (`supercells`), coordinates of the supercell centers (`x` and `y`), and an average of all of the input variables (`elevation` in this case).

It allows us to plot the supercells' borders on top of the input raster data:

```{r}
plot(vol)
plot(st_geometry(vol_slic1), add = TRUE, lwd = 0.2)
```

The `vol_slic1` object can be a final result of the analysis or used for further grouping or clustering.

# Customized arguments

The `supercells()` function also has a set of optional arguments, including:

- `step` - a distance (in the number of cells) between initial superpixel' centers. It can be used instead of `k`
- `dist_fun` - a distance function used to calculate similarities between raster values 
- `clean` - should connectivity of the supercell be enforced? If `FALSE`, then some supercells could consist of several nonadjacent regions. It is `TRUE` by default.
- `iter` - a number of iterations performed to create the output. It is `10` by default.

Let's try to change the values of some of them:

```{r, results='hide'}
vol_slic2 = supercells(vol, step = 11, compactness = 1, dist_fun = "euclidean", 
                       clean = FALSE, iter = 10)
```

The output still has the same structure (is an `sf` object with supercell ids, coordinates of their centers, and an average of all of the input variables), but has a slightly different number of supercells, and their borders are a bit different.

```{r}
plot(vol)
plot(st_geometry(vol_slic2), add = TRUE, lwd = 0.2)
```

# References
