---
title: "Supercells of spatial categorical patterns"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supercells of spatial categorical patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r system.file('refs.bib', package = 'supercells')`"
nocite: '@*'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4
)
```

> **Version note:** This vignette documents the `supercells()` interface as it existed in version 1.0 of the package. Some arguments, defaults, and behaviors may differ in newer releases. For up-to-date details, see `?sc_slic` and the current reference docs.

Superpixels are a collection of segmentation concepts of grouping pixels with similar characteristics.
In this package, we refer to them as supercells.
It is often used in computer vision to delineate parts of RGB images that are more meaningful and easier to analyze.
When applied to RGB images, each superpixel contains similar colors that also could represent real-world objects.
A large number of methods for creating superpixels were developed in the last decades, with the SLIC algorithm (Achanta et al. (2012), <doi:10.1109/TPAMI.2012.120>) being the most prominent.

The **supercells** package aims to utilize the concept of supercells for a variety of spatial data. 
This package works on spatial data with one variable (e.g., continuous raster), many variables (e.g., RGB rasters), and spatial patterns (e.g., areas in categorical rasters).
Therefore, it enables not only to find areas that look similar on an RGB (satellite) image, but also to regionalize areas with comparable values of one or more variables.

This vignette shows how to use the **supercells** package on a categorical raster dataset.
To reproduce the following results on your own computer, install and attach the packages:

```{r}
library(supercells)    # supercells for spatial data
library(terra)         # spatial raster data reading and handling
library(sf)            # spatial vector data reading and handling
library(motif)         # tools for working with categorical spatial patterns
```

The first step is to read the input data.
This time our input data `landcover2015.tif`, included in the **motif** package, that contains land cover data for New Guinea.

```{r}
landcover = rast(system.file("raster/landcover2015.tif", package = "motif"))
plot(landcover)
```

Next, we need to derive some key information about the categorical spatial patterns of the New Guinea land cover.
This can be done with **motif** -- below, we calculate compositions of land cover categories (learn about it more in @nowosad_motif_2021) for non-overlapping windows of 20 by 20 cells, and then convert it back into a **terra** object.

```{r, message = FALSE}
comp_output = lsp_signature(landcover, type = "composition", window = 20,
                             normalization = "pdf", ordered = FALSE)
#comp_output = lsp_restructure(comp_output)
comp_output = lsp_add_terra(comp_output)
comp_output
```

The result, `comp_output`, contains two layers with metadata (`id` and `na_prop`), and sever layers with a proportion of each land cover category for each window.
We can select only the latter layers as the metadata is not needed for the following calculations.

```{r}
comp_output2 = subset(comp_output, 3:9)
plot(comp_output2)
```

Finally, we are able to use the `supercells()` function to delineate areas with similar proportions of different land cover categories.
I decided to derive about 2,000 supercells that will be created based on the Jensen-Shannon distance (`dist_fun = "jsd"`).
This distance measure is more suitable for proportions (e.g., 0-1) than the default Euclidean distance.

```{r, results='hide', eval=TRUE}
slic = supercells(comp_output2, k = 2000, compactness = 0.1, dist_fun = "jsd")
```

The result is an `sf` object containing derived supercells with attributes storing the average proportions for each supercell:

```{r}
plot(landcover, reset = FALSE)
plot(st_geometry(slic), add = TRUE, border = "red")
```

# References
