---
title: "Introduction to supercells"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to supercells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r system.file('refs.bib', package = 'supercells')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **supercells** package creates compact, homogeneous regions from raster data.
It replaces thousands or millions of pixels with a smaller number of meaningful spatial units.
This helps with interpretation, visualization, and downstream modeling.
It also reduces noise while preserving important spatial structure.

The package is designed to work with rasters that have many layers or time steps.
It supports multiple distance measures such as Jensen–Shannon divergence or dynamic time warping, as well as user-defined distance functions, so similarity can be adapted to the data.
It can therefore handle diverse data types and high-dimensional inputs without changing the main workflow.
The package core ideas are described in @Nowosad2022.

# Core functions

The main function in the package is `sc_slic()`.
It implements a SLIC-style algorithm that balances spatial compactness with value similarity [@achanta2012slic] and extends it to work with multi-layer rasters and custom distance functions.
The output is a set of polygon supercells stored as an `sf` object.
This object can include centers, average values, and IDs depending on the requested outcomes.
In practice, `sc_slic()` is the entry point for most workflows.

The key tuning arguments in `sc_slic()` are `step` and `compactness`.
`step` sets the spacing of initial centers and controls the expected number of supercells.
It can be thought of as the expected spatial scale of the output. 
`compactness` controls the tradeoff between spatial regularity and value similarity.
Lower values prioritize value similarity and may lead to irregular shapes, while higher values prioritize shape regularity -- supercells may look more like squares -- but may be less homogeneous in terms of values.

For compactness selection, you may use `sc_tune_compactness()` to estimate a good value from a short pilot run.
Alternatively, the `"auto"` option for `compactness` enables SLIC0-style adaptive compactness when you want the algorithm to adjust locally -- this does not require setting a specific value, but also takes away direct control.

To assess quality of the resulting supercells, use `sc_metrics_pixels()` for pixel-level distances, `sc_metrics_supercells()` for per-region summaries, and `sc_metrics_global()` for a general overview.
These metrics help compare different parameter settings or input preprocessing choices.

# Minimal example

The goal of this example is to derive supercells from a raster and visualize them.
For that purpose, we will use the built-in `volcano` dataset, which is a small raster of elevation values.

```{r}
#| message: false
library(supercells)
library(terra)

vol <- terra::rast(system.file("raster/volcano.tif", package = "supercells"))
```

This raster has `r terra::ncell(vol)` cells -- a small number for demonstration, but the same workflow applies to much larger rasters as well.
Moreover, it has only one layer -- but the same functions can handle multi-layer rasters with any number of layers.

Below, we create supercells in three different formats: polygons, points, and raster IDs.
The polygon supercells are the most common output.
Here, we specified `step = 8` which would result in supercells of approximately 8x8 cells in size if they were perfectly regular, but the actual shapes and sizes will depend on the data and the `compactness` setting.
The compactness is set to `1` -- the behavior of this parameter depends on many factors, including the range of values in the raster, their properties, and the selected distance measure.

```{r}
# Polygon supercells (sf)
vol_sc <- sc_slic(
  vol,
  step = 8,
  compactness = 1
)

terra::plot(vol)
plot(sf::st_geometry(vol_sc), add = TRUE, lwd = 0.6, border = "red")
```

The resulting `sf` object contains one row per supercell.
Each row stores summary values of each layer in the original raster, as well as the geometry of the supercell.
By default only summary values are returned, so use `outcomes = c("supercells", "coordinates", "values")` when you also want IDs and center coordinates.

Two related functions provide alternative output formats.
Use `sc_slic_points()` to return only supercell centers as points.
Use `sc_slic_raster()` to return a raster of supercell IDs for large datasets or raster-based pipelines.
These functions share the same core arguments and can be swapped with minimal changes.

Now, let's try to tune the `compactness` parameter using a pilot run.

```{r}
tune <- sc_tune_compactness(
  vol,
  step = 8,
  metrics = "local"
)
tune
```

Next, we may try to create supercells with the suggested `compactness` value.

```{r}
vol_sc_tuned <- sc_slic(
  vol,
  step = 8,
  compactness = tune$compactness_local
)

terra::plot(vol)
plot(sf::st_geometry(vol_sc_tuned), add = TRUE, lwd = 0.6, border = "red")
```

Such a derived compactness value may serve as a good starting point for further experimentation.

Afterwards, we can evaluate the quality of the resulting supercells using metrics on different levels.

```{r}
pixel_metrics <- sc_metrics_pixels(vol, vol_sc_tuned)
supercell_metrics <- sc_metrics_supercells(vol, vol_sc_tuned)
global_metrics <- sc_metrics_global(vol, vol_sc_tuned)
```

The pixel-level metrics show how well the supercells represent the original raster values at the pixel level.
By default they include `spatial`, `value`, and `combined` distances, with `balance` available as an optional fourth metric.
When `scale = TRUE` (the default), the spatial and value layers are returned as `spatial_scaled` and `value_scaled`.

```{r}
plot(pixel_metrics, nr = 1)
```

These maps help diagnose where spatial or value coherence is weak.
For example, large patches of high values can indicate areas where the chosen parameters may be too coarse for local variation.

The supercell-level metrics summarize each polygon as a single value.
These are helpful when you want to compare regions or join metrics to attributes.

```{r}
plot(supercell_metrics)
```

The global metrics provide a single-row summary of overall quality.
They are useful for comparing parameter settings across multiple runs, such as a small grid of `step` and `compactness` values.

```{r}
global_metrics
```

<!-- 
# Where to go next

See `v2-workflows` for end-to-end usage patterns and practical recipes.
See `v2-tuning` for guidance on choosing `compactness` and related settings.
See `v2-evaluation` for metrics and diagnostics that help compare results.
See `v2-benchmarks` for performance notes and scaling behavior.
See `v2-changes-since-v1` for a concise summary of recent updates. -->
