---
title: "Choosing parameters for supercells"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing parameters for supercells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette focuses on how parameters affect the supercell output, with the main choices being `step` or `k`, the `compactness` value, and the distance and averaging functions.
The goal is not to find one universal set of parameters, but to choose values that match your intent.

All examples use the `volcano` raster for simplicity, but the same ideas apply to larger and multi-layer rasters.

```{r}
#| message: false
library(supercells)
library(terra)

vol <- terra::rast(system.file("raster/volcano.tif", package = "supercells"))
```

# Supercells algorithm

The supercells algorithm is an iterative process that starts with initial centers and assigns pixels to the nearest center based on a combined distance that incorporates both value similarity and spatial proximity.
After the initial assignment, the algorithm updates the centers by averaging the values of the assigned pixels and recalculating the combined distance for the next iteration.
This process continues until the specified number of iterations is reached.

In this package, the combined distance follows the standard SLIC form:

$$
D = \sqrt{\left(\frac{d_s}{\text{step}}\right)^2 + \left(\frac{d_v}{c}\right)^2}
$$

where $d_s$ is the spatial distance in grid-cell units, $d_v$ is the value-space distance (from `dist_fun`), and $c$ is the `compactness` value.
When `step_unit = "map"`, the `step` value is converted to cells before segmentation; distances are still computed in grid-cell units.
Larger `compactness` down-weights the value term, making shapes more regular, while smaller values emphasize value similarity.

# Choosing step or k

You can control the number and size of supercells using either `step` or `k`.
`step` defines the spacing of initial centers in pixel units (or map units when `step_unit = "map"`).
Smaller `step` values produce more and smaller supercells, while larger values produce fewer and larger supercells.
For example, `step = 8` creates centers approximately every 8 pixels, which leads to supercells that are roughly 8 by 8 pixels in size, depending on the compactness.

```{r}
sc_step <- sc_slic(vol, step = 8, compactness = 5)
```

By default, `step` is in pixel units, but instead we can also specify it in map units with `step_unit = "map"`.
This allows us to think about the spatial scales of expected supercells in terms of the actual map units.
For example, if your raster has a resolution of 10 meters, then `step = 200` with `step_unit = "map"` would create centers approximately every 200 meters, which corresponds to every 20 pixels.

```{r}
sc_step_map <- sc_slic(vol, step = 200, step_unit = "map", compactness = 5)
```

`k` specifies the desired number of supercells and the algorithm chooses an approximate step size.
This is convenient when you want a target number instead of a target spatial scale.

```{r}
sc_k <- sc_slic(vol, k = 100, compactness = 5)
```

In practice, use `step` when spatial scale matters, and use `k` when you want a fixed count across datasets.
Importantly, both approaches still require a sensible `compactness` value.

<!--
1.
You can also preview the initial center spacing with `sc_slic_points(iter = 0)`.
This can help confirm whether your `step` or `k` choice aligns with your intended scale.

```{r}
centers_preview <- sc_slic_points(vol, step = 8, compactness = 5, iter = 0)
terra::plot(vol)
plot(sf::st_geometry(centers_preview), add = TRUE, pch = 3, col = "red")
```

2. also custom centers
-->

```{r}
#| fig-show: "hold"
sc_step_small <- sc_slic(vol, step = 6, compactness = 5)
sc_step_large <- sc_slic(vol, step = 12, compactness = 5)
terra::plot(vol, main = "step = 6")
plot(sc_step_small[0], add = TRUE, border = "red", lwd = 0.5)
terra::plot(vol, main = "step = 12")
plot(sc_step_large[0], add = TRUE, border = "red", lwd = 0.5)
```

# Choosing compactness

`compactness` controls the tradeoff between spatial regularity and value similarity:
lower values favor value similarity and may create irregular shapes, while higher values favor more regular shapes but can mix dissimilar values.

In general, it is hard to predict the best `compactness` value without testing, and it depends on many factors including the range and distribution of values in your raster, the selected distance function, and the spatial structure of the data.

You can see this tradeoff by comparing two compactness values with the same step.

```{r}
#| fig-show: "hold"
sc_compact_low <- sc_slic(vol, step = 8, compactness = 1)
sc_compact_high <- sc_slic(vol, step = 8, compactness = 10)
terra::plot(vol, main = "compactness = 1")
plot(sc_compact_low[0], add = TRUE, border = "red", lwd = 0.5)
terra::plot(vol, main = "compactness = 10")
plot(sc_compact_high[0], add = TRUE, border = "red", lwd = 0.5)
```

## Tuning compactness

The `sc_tune_compactness()` function estimates a reasonable starting value from a short run of the algorithm.

It supports two summaries with `metrics = "global"` and `metrics = "local"`.
The global version looks at overall balance between value and spatial distances, while the local version uses a neighborhood-based value scale.
More precisely:

- **Global**: runs a short pilot segmentation (`iter = 1` by default), computes pixel-level `spatial` and `value` distances, then takes their medians over pixels. The compactness is estimated as `compactness = (median(value) / value_scale) * step / median(spatial)`.  
  This aligns the median value and spatial terms in the combined distance.
- **Local**: computes, for each center, the mean value distance within a local $2 \times \text{step}$ window, then returns the median of those per-center means (after `value_scale`).  
  This yields a compactness tied to local value variability, without explicitly using spatial distances.

The local estimate is often more stable for heterogeneous rasters.

```{r}
tune_global <- sc_tune_compactness(vol, step = 8, metrics = "global")
tune_local <- sc_tune_compactness(vol, step = 8, metrics = "local")
tune_global
tune_local
```

Both results return a one-row data frame with `step`, `metric`, and `compactness`.
You can plug the suggested value into `sc_slic()` by setting `compactness` to the estimated value.

```{r}
sc_tuned <- sc_slic(vol, step = 8, compactness = tune_global$compactness)
```

If your raster has many layers, the value distances can be large.
The `value_scale` argument controls the scaling of value distances before the compactness estimate.
The default `"auto"` divides by `sqrt(nlyr(x))`, which is often a good baseline for high-dimensional inputs.
If your values are already standardized or on a common scale, you can set `value_scale = 1`.

## Automatic compactness

For heterogeneous rasters, `compactness = "auto"` enables SLIC0-style adaptive compactness.
This adjusts the value scale per supercell and often improves local adaptation.
At the same time, it reduces direct control, so use it when a single global compactness is hard to choose.
Importantly, it still uses your chosen `dist_fun` for value distances.

```{r}
sc_auto <- sc_slic(vol, step = 8, compactness = "auto")
```

```{r}
terra::plot(vol, main = "compactness = auto")
plot(sc_auto[0], add = TRUE, border = "red", lwd = 0.5)
```

When you compare results, keep in mind that metrics from `"auto"` are not directly comparable to fixed-compactness runs.
You can still compare spatial metrics, but value metrics reflect local scaling when `compactness = "auto"`.

## Manual tuning

A parallel approach to deciding on `compactness` is to run a small grid of values and compare the global metrics.
It allows you to see the tradeoffs between spatial and value distances across a range of compactness settings.

```{r}
cmp_vals <- c(1, 3, 5, 8)
cmp_metrics <- lapply(cmp_vals, function(cmp) {
  sc_i <- sc_slic(vol, step = 8, compactness = cmp)
  sc_metrics_global(vol, sc_i, scale = TRUE)
})
cmp_metrics <- do.call(rbind, cmp_metrics)
cmp_metrics
```

# Distance and averaging functions

The `dist_fun` argument defines how similarity is measured in value space.
Built-in options include `"euclidean"`, Jensen–Shannon divergence, and dynamic time warping.
You can also pass a name of the distance measure from the **philentropy** package^[
See the **philentropy** documentation at <https://drostlab.github.io/philentropy/articles/Distances.html>.] or a custom function that takes two numeric vectors and returns a single number.
If you use a custom distance, ensure it returns a single finite number.

```{r}
sc_manhattan <- sc_slic(vol, step = 8, compactness = 5, dist_fun = "manhattan")
```

The `avg_fun` argument controls how values are summarized within each supercell.
The SLIC algorithm iteratively updates supercell centers based on the average value of the assigned pixels -- and with this argument you can choose how that average is calculated.
The default is `"mean"`, and `"median"` is an alternative, often more robust to outliers.

```{r}
sc_median <- sc_slic(vol, step = 8, compactness = 5, avg_fun = "median")
```

```{r}
#| fig-show: "hold"
terra::plot(vol, main = "dist_fun = manhattan")
plot(sf::st_geometry(sc_manhattan), add = TRUE, border = "red", lwd = 0.5)
terra::plot(vol, main = "avg_fun = median")
plot(sf::st_geometry(sc_median), add = TRUE, border = "red", lwd = 0.5)
```

If you change `dist_fun` or `avg_fun`, you should remember to update `compactness`.
Different distances change the scale of value differences and therefore the balance with the spatial term.

After tuning, use `sc_metrics_pixels()`, `sc_metrics_supercells()`, or `sc_metrics_global()` to compare runs.
These functions accept the raster as `x` and the supercells as `sc`, and they reuse the stored `dist_fun` when available.
See the evaluation vignette for guidance on interpreting these metrics: <https://jakubnowosad.com/supercells/articles/v2-evaluation.html>.

# Connectivity and minimum size

The `clean` argument, enabled by default, enforces connectivity and removes small fragments that may arise during the iterative process of assigning pixels to supercells.
This is usually desirable for polygon outputs and downstream analyses.
The `minarea` argument sets a minimum expected size in pixels for supercells, and smaller supercells are merged into neighbors.
By default, when `clean = TRUE`, a minimum size is still enforced internally (about one quarter of the average supercell size), so small fragments are merged unless you turn off cleaning.
If you see many tiny supercells, increase `minarea` or `step`.

```{r}
sc_clean <- sc_slic(vol, step = 8, compactness = 5, clean = TRUE, minarea = 20)
```

If you need unpostprocessed supercells for speed or debugging, you can set `clean = FALSE`.

# Iterations

You can inspect how the algorithm converges over iterations by enabling `iter_diagnostics = TRUE`.
This attaches diagnostics to the output and allows plotting with `sc_plot_iter_diagnostics()`.^[It only works when chunking is disabled.]

```{r}
sc_diag <- sc_slic(vol, step = 8, compactness = 5, iter_diagnostics = TRUE)
sc_plot_iter_diagnostics(sc_diag)
```

Use the diagnostics to decide whether fewer iterations are sufficient.
If the curves stabilize early, you can reduce `iter` to speed up large runs.
